Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2022-12-15 manuscript.Rmd
Added discussion, began writing abstract and conclusion.

@@

---------------------------------
OK, some pointers for the introduction:

I think we need to draw a distinction between binocular summation at threshold (on which there is a big literature) and binocular facilitation above threshold. Much of the introduction is focussed on threshold measures, but actually I think our focus here is more about suprathreshold conditions where we usually see little facilitation. I think it will help the reader if we use the term summation only to refer to threshold phenomena, and facilitation in the more general case (I've tried to do this elsewhere in the paper).

The key points should be:

1. For grating stimuli, there are summation effects at threshold, but these are lost at high contrasts.

2. Binocular combination happens in multiple pathways, including the pupil pathway. (This is a good place to mention the Quaia paper).

3. We don't know much about the pupil response, though it is clearly binocular because of the consensual response (convergence and divergence seem to be other ways of saying the same thing, so not sure they are worth mentioning).

4. Actually there's not much work on temporal binocular combination even in the primary visual pathway.

5. We have developed a novel paradigm that allows us to probe both cortical and subcortical pathways simultaneously for flickering light.

I wouldn't say too much about models, or about the matching experiment, at this stage.
---------------------------------

Binocular combination provides a higher visual sensitivity than monocular viewing. This superiority is known as binocular summation and is defined by the binocular summation ratio (BSR), which was originally believed to be around $\sqrt{2}$ ($\approx1.4$) for grating stimuli at detection threshold [@Campbell1965]. In other words, a monocular stimulus can elicit the same response as a binocular stimulus if it has a contrast that is 1.4 times higher. This led Legge to develop a widely accepted explanation that used quadratic summation to describe binocular combination [@Legge1984]: monocular signals from the right (R) and the left (L) eyes are squared before being summed together and the binocular response (B) is given by the square root of the output (B = $\sqrt{R^2 + L^2}$, when R and L are equal to 1, the output is $\sqrt{2}$). However, subsequent research has shown that these explanations are not fully adequate to account for binocular summation. Both of these accounts constitute single channel models, and this type of model has been shown to not being able to account for contrast detection in the presence of noise [@Anderson1989]. Moreover, more recent research has shown that the summation ratio can vary greatly between $\sqrt{2}$ and 2 depending on factors such as the spatial and temporal frequency of a stimulus or the sensitivity difference between the eyes [@Baker2018]. These observations led to the development of multistage gain control models, which combine binocular summation and interocular suppression, and can account for contrast matching, detection and discrimination for spatial contrast [@Ding2006; @Meese2006].

In general, it seems that the mechanisms behind binocular combination have been thoroughly studied, as have been the anatomical pathways behind it: light enters the eye though the pupil and signals are sent from the left and right retinae to the primary visual cortex, remaining anatomically isolated while passing through the lateral geniculate nucleus (LGN) until they reach V1, where they are binocularly combined [@Purves2008]. However, there is an eye component that is often underestimated in its role to determine the quality of visual information: the pupil. The pupils are openings found in the centre of the eyes that appear to be black and allow light to enter the eyes. Their size determines how much light will reach the retina and it is usually determined by the ambient levels of light: in brightness the pupils will constrict and in darkness they will dilate. This is known as the pupillary light response (PLR). The anatomical pathways that regulate this response are well understood and are very clearly and extensively described in the literature [@Angee2021; @Mathot2018; @McDougal2010; @Wang2015]. However, they are anatomically distinct from the LGN-V1 pathway meaning that binocular combination occurs separately in anatomically distinct pathways. Given this, not much is known about the computational processes behind the PLR except for some evidence of binocular interaction. The presence of a consensual response in one eye when the other is being stimulated, and the presence of convergence (one pupil responds to illumination in either retina) and divergence (both pupils respond to illumination of one retina) [@Wyatt1981] are evidence of this binocular interaction.

With this in mind we designed an experiment that simultaneously recorded electrophysiological and pupillometric responses to investigate the combination of flickering light signals in both the visual cortex and the pupils. The results should offer new insight about basic neural circuits and information on how they might be affected in clinical disorders of vision (e.g. amblyopia). Based on previous literature, we expected to find a non-linear combination of the responses in visual cortex, as described by the gain control mechanisms, and a more linear combination of the responses and a greater binocular response at the level of the pupils.

Several models have been developed and proposed to describe binocular combination and contrast matching results for spatial contrast. Among these, the models proposed by Ding and Sperling (2006) and Meese and colleagues (2006) are widely accepted. These models incorporate a dynamic contrast gain control, which allows to develop a model that combines binocular summation and interocular suppression The first model is a multistage model that consists of two pairs of gain control mechanisms that inhibit each other [@Ding2006]. In the earlier stage, the right and left eye channels exert gain control on the other channel. Then, they exert gain control on the other channel’s gain control. Lastly, the outputs from each channel are summed binocularly to determine the binocular signal’s magnitude. The second model is similar as it also incorporates gain control mechanisms [@Meese2006]. In the first stage, it is applied to the left and right eye channels with suppression from the other eye. Then the signal from the two eyes is summed binocularly and gain control is applied a second time. This second model in particular is interesting and relevant as it was designed to also explain data from matching experiments.

To follow up on the results that we obtained, we decided to perform a contrast matching experiment to investigate whether perception of flickering light is consistent with the results observed in the cortical pathways (visual cortex) or the subcortical pathways (pupils). Matching is a paradigm in which the perceived brightness of a standard stimulus is matched to that of a target stimulus. In the latter, the interocular ratios of luminance are varied to obtain an equibrightness curve. Previous literature has used this paradigm to investigate the binocular fusion of static stimuli and the temporal combination of spatial flickering of spatial increments (a bright target on a dark background) and decrements (a dark target on a bright background) [@Anstis1998; @Levelt1965]. For spatial increments, it was found that binocular fusion seems to follow approximately linear combination rules. This means that, for a monocular stimulus to elicit the same response as a binocular stimulus, the former needs to have twice the signal of the latter. On the other hand, spatial decrements follow a winner-takes-all pattern. This means that the observer is seeing what the eye that is receiving the strongest signal is seeing.

Our experiment used different stimuli than the ones used by @Anstis1998: in the binocular fusion experiment, their stimuli were not flickering and, in the flicker experiment, they were always shown in both eyes. In our experiment, in some conditions, the flicker was shown to only one eye. Moreover, they were looking at the temporal fusion of the flicker while we focussed on the binocular fusion of the flicker. Based on this and on the results from our first experiment, we expected to find a near linear summation of the responses.

@@

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The fast Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit sine-waves to the data, rather than using Fourier analysis [e.g. @Spitschan2014]. Our novel approach is more robust to noise at other frequencies, and makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [e.g. @Figueira2022]

To analyse the matching data, we pooled the trial responses across all repetitions of a given condition for each participant. We then fitted a cumulative normal psychometric function to estimate the point of subjective equality at the 50\% level. Thresholds were averaged across participants in logarithmic (dB) units. For all experiments, we used a bootstrapping procedure with `r nbootstraps` iterations to estimate standard errors across participants. All analysis and figure construction was conducted using a single R-script, available online, making this study fully computationally reproducible.

@@

Results from the temporal frequency experiment are shown in \@ref(fig:TFdata). Figure \@ref(fig:TFdata)a shows the Fourier spectra for responses to binocular flicker at 5 different frequencies (2, 4, 8, 16, and 30 Hz). From 2 to 16 Hz, clear signals are observed at each fundamental frequency, and typically also their higher harmonics (integer multiples of the fundamental). However, at 30 Hz (upper row), the responses recorded were not demonstrably above the noise baseline. Figure \@ref(fig:TFdata)b compares the monocular and binocular responses at each stimulation frequency. Here we replicate the substantial summation effect across frequencies up to and including 16Hz (Fig. \@ref(fig:TFdata)c), demonstrating that strong binocular facilitation in the EEG data of Experiment 1 cannot be attributed to our use of 2Hz flicker.

@@

>>>>

We investigated the rules of binocular combination of temporal flickering stimuli. Using EEG and pupillometry simultaneously, we investigated the combination of light discs at five temporal contrast levels flickering at 2 Hz in the cortical and subcortical pathways. Results showed a strong binocular facilitation in the visual cortex and a non-linear combination process with evidence of suppression in the subcortical pathways. Following these findings, we performed a matching experiment using the same type of stimuli and found equivalent linear summation effects. Using a Bayesian approach, all our data were fitted to a computational model that showed a stronger interocular suppression for the pupillometry data compared with the EEG and matching data.

---
OK, some pointers for the introduction: I think we need to draw a distinction between binocular summation at threshold (on which there is a big literature) and binocular facilitation above threshold. Much of the introduction is focussed on threshold measures, but actually I think our focus here is more about suprathreshold conditions where we usually see little facilitation. I think it will help the reader if we use the term summation only to refer to threshold phenomena, and facilitation in the more general case (I’ve tried to do this elsewhere in the paper).
The key points should be:
1. For grating stimuli, there are summation effects at threshold, but these are lost at high contrasts.
2. Binocular combination happens in multiple pathways, including the pupil pathway. (This is a good place to mention the Quaia paper).
3. We don’t know much about the pupil response, though it is clearly binocular because of the consensual response (convergence and divergence seem to be other ways of saying the same thing, so not sure they are worth mentioning).
4. Actually there’s not much work on temporal binocular combination even in the primary visual
pathway.
5. We have developed a novel paradigm that allows us to probe both cortical and subcortical pathways simultaneously for flickering light.
I wouldn’t say too much about models, or about the matching experiment, at this stage.

@@

Binocular combination provides a higher visual sensitivity than monocular viewing. At threshold, this superiority is known as binocular summation, while, above threshold, we refer to this as binocular facilitation. Binocular summation is defined by the binocular summation ratio (BSR). This was originally believed to be around $\sqrt{2}$ ($\approx1.4$) for grating stimuli at detection threshold [@Campbell1965]. In other words, a monocular grating can elicit the same response as a binocular grating if it has a contrast that is 1.4 times higher. This led Legge to develop a widely accepted explanation that used quadratic summation to describe binocular combination for grating stimuli [@Legge1984]: monocular signals from the right (R) and the left (L) eyes are squared before being summed together and the binocular response (B) is given by the square root of the output (B = $\sqrt{R^2 + L^2}$, when R and L are equal to 1, the output is $\sqrt{2}$). However, subsequent research has shown that these explanations are not fully adequate to account for binocular facilitation above threshold, as this summation effects are lost for gratings at high contrast. Both of these accounts constitute single channel models, and this type of model has been shown to not being able to account for contrast detection in the presence of noise [@Anderson1989]. Moreover, more recent research has shown that the summation ratio can vary greatly between $\sqrt{2}$ and 2 depending on factors such as the spatial and temporal frequency of a stimulus or the sensitivity difference between the eyes [@Baker2018]. These observations led to the development of multistage gain control models, which combine binocular summation and interocular suppression, and can account for contrast matching, detection and discrimination for spatial contrast [@Ding2006; @Meese2006].

In general, it seems that the mechanisms behind spatial binocular combination have been thoroughly studied, as have been the anatomical pathways behind it: light enters the eye though the pupil and signals are sent from the left and right retinae to the primary visual cortex, remaining anatomically isolated while passing through the lateral geniculate nucleus (LGN) until they reach V1, where they are binocularly combined [@Purves2008]. However, there is an eye component that is often underestimated in its role to determine the quality of visual information: the pupil. The pupils are openings found in the centre of the eyes that appear to be black and allow light to enter the eyes. Their size determines how much light will reach the retina and it is usually determined by the ambient levels of light: in brightness the pupils will constrict and in darkness they will dilate. This is known as the pupillary light response (PLR). The anatomical pathways that regulate this response are well understood and are very clearly and extensively described in the literature [@Angee2021; @Mathot2018; @McDougal2010; @Wang2015]. However, they are anatomically distinct from the LGN-V1 pathway meaning that binocular combination occurs separately in anatomically distinct pathways. Given this, not much is known about the computational processes behind the PLR although there is some evidence of binocular interaction and binocular combination happening at the level of the pupil pathways. Evidence of binocular interaction is provided by the presence of a consensual response in the pupils: when only one pupil is stimulated, an equivalent response will be recorded in the unstimulated pupil [@Wyatt1981]. Evidence of binocular combination is provided by the binocular facilitation observed for the ocular following eye movements in response to fast moving stimuli: when comparing the amplitude of the responses to monocular and binocular stimuli, the binocular response will be more than twice the monocular response [@Quaia2018].

With this in mind, we designed an experiment that simultaneously recorded electrophysiological and pupillometric responses using changes in the temporal contrast of a flickering light. This is a novel paradigm that we developed that allowed us to probe both cortical (using EEG) and subcortical (using a binocular eyetracker) pathways simultaneously for flickering light. The results should offer new insight about basic neural circuits and information on how they might be affected in clinical disorders of vision (e.g. amblyopia). Based on previous literature, we expected to find a non-linear combination of the responses in the visual cortex, as described by the gain control mechanisms used to describe spatial contrast combination, and a more linear combination of the responses and a greater binocular response at the level of the pupils. Additionally, we decided to follow up on our results from this first experiments to test whether perception of flickering light is consistent with the results that we observed in the cortical pathways (visual cortex). We therefore conducted a contrast matching experiment using the same type of flickering stimuli used in our first experiment.

@@

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The fast Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit sine-waves to the data, rather than using Fourier analysis [e.g. @Spitschan2014]. The Fourier approach is more robust to noise at other frequencies and has already been used in some previous studies [see @Barrionuevo2014; @Barrionuevo2016]. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis [e.g. @Figueira2022]

To analyse the matching data, we pooled the trial responses across all repetitions of a given condition for each participant. We then fitted a cumulative normal psychometric function to estimate the point of subjective equality at the 50\% level. Thresholds were averaged across participants in logarithmic (dB) units.

For all experiments, we used a bootstrapping procedure with `r nbootstraps` iterations to estimate standard errors across participants. All analysis and figure construction was conducted using a single R-script, available online, making this study fully computationally reproducible.

@@

Results from the temporal frequency experiment are shown in figure \@ref(fig:TFdata). Figure \@ref(fig:TFdata)a shows the Fourier spectra for responses to binocular flicker at 5 different frequencies (2, 4, 8, 16, and 30 Hz). From 2 to 16 Hz, clear signals are observed at each fundamental frequency, and typically also their higher harmonics (integer multiples of the fundamental). However, at 30 Hz (upper row), the responses recorded were not demonstrably above the noise baseline. Figure \@ref(fig:TFdata)b compares the monocular and binocular responses at each stimulation frequency. Here we replicate the substantial summation effect across frequencies up to and including 16Hz (Fig. \@ref(fig:TFdata)c), demonstrating that strong binocular facilitation in the EEG data of Experiment 1 cannot be attributed to our use of 2Hz flicker.

@@

Using a novel paradigm that combines elctrophysiological and pupillometic measures, we found surprising results for binocular combination of flickering light: in the visual cortex, the processing of temporal luminance modulations seems to happen linearly, with no evidence of interocular suppression, while, in the subcortical pathways, this seems to happen non-linearly with evidence of interocular suppression. This pattern of results was also supported by the computational model that we fitted our data to, which showed a much greater suppression weight of the pupillometry data compared with the EEG data. Additionally, we also found that perception of flickering light is consistent with cortical responses, with our matching results also showing linear summation effects.

The results from our first experiment were unexpected for both the pupillometry and the EEG. Previous studies investigating binocular combination of grating stimuli using steady-state EEG showed that combination for this type of stimuli seems to happen in a non-linear way [@Baker2017]. However, we found that there seems to be a strong binocular facilitation for flickering stimuli. While this could have been due to the lower frequency used in our experiment, our second experiment showed that this was not the case. A reason for this apparent facilitation could be due to the different nature of our stimuli: @Baker2017 used static gratings, while here we used flickering stimuli. @Quaia2018 observed a strong binocular facilitation in eye movement response to moving stimuli. Here, obviously, we were not investigating eye movements, but our stimuli included a movement component due to their flickering nature. This induced movement of the stimuli could be the reason why we observed this strong facilitation in the visual cortex. As for our pupillometry results, we observed less binocular facilitation than what we hypothesised at beginning of testing. As mentioned in the introduction, the presence of a consensual response in the pupils [@Wyatt1981] is evidence of binocular interaction. @Wyatt1981 measured an inequality of the responses between the two eyes during this consensual response, which could actually be explained by the presence of interocular suppression between the two pupils.

On the other hand, the results of our matching experiment were not completely unexpected. Previous literature has used the matching paradigm to investigate the binocular fusion of static stimuli for spatial increments, which consists of a bright target on a dark background, and spatial decrements, which consist of a dark target on a bright background [@Anstis1998; @Levelt1965]. For the former, it was found that binocular fusion seems to follow approximately linear combination rules, while, for the latter, it was found that binocular fusion seems to follow a winner-take-all pattern, where the observer is seeing what the eye that is receiving the strongest signal is seeing. Our flickering stimuli were a combination of both increments and decrements presented on a dark background, but presented a stronger similarity in presentation to the increments than the decrements. This would explain why our results showed a strong facilitation.

As mentioned before, the main frequency for the three experiments was 2 Hz. The reason behind this was that we wanted to be able to measure both EEG and pupillary responses simultaneously. However, these two methods have different temporal properties: pupillometry is most responsive to slow flicker frequencies of less than 1 Hz [@Spitschan2014], while steady-state EEG generates stronger responses to faster frequencies of more than 3 Hz [@Norcia2015]. This means that stronger responses for pupillometry should be obtained at a lower frequency than the one we used in the main experiment, and for EEG at a higher frequency, meaning that our interpretation of the results might have been limited by the frequency that we selected. However, previous studies have shown that significant pupil responses can be obtained at higher frequencies, with the pupils showing a similar pattern of responses as at 1 Hz or lower, although at a reduced amplitude [@Barrionuevo2014; @Barrionuevo2016]. Additionally, our second experiment shows that the strong binocular facilitation observed in the cortical areas for flickering stimuli at 2 Hz is also observable at higher frequencies. Therefore, this would seem to indicate that 2 Hz was a good compromise to obtain significant responses from both pupillometry and EEG.

# Conclusions

With these three experiments, we have demonstrated that binocular combination for temporal flickering stimuli seems to happen in different manners in cortical and subcortical pathways.
